PANCHANGE_data_preprocessing

### 2020-08-07 PANCHANGE_data_preprocessing2020-07-15.Rmd
admin changes
updated the demographics read in and changed Age_raw to Age_uncleaned

# Select variables
```{r glad dem cleaning}
glad.dem.raw.id.unclean <- glad.dem.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
         Sample, # Sample
         ID, # ID
         startDate.glad, #Date when participant signed up to glad
         Age_raw = dem.how_old_are_you_now, # Age
         Gender = dem.which_gender_do_you_identify_with, # Gender
         Sex = dem.select_questionnaire_items_medical, #Sex
         EduYrs = dem.school_education_school_years, # Years in education
         dem.college_or_university_degree_numeric, # level of education - univeristy
         dem.a_levelsas_levels_or_equivalent_numeric, # level of education - alevel
         dem.o_levelsgcses_or_equivalent_numeric, # level of education - gcse
         dem.cses_or_equivalent_numeric, # level of education - cse
         dem.nvq_or_hnd_or_hnc_or_equivalent_numeric, # level of education - nvq
         Ethnicity = dem.questions_based_ethnic_origin # Ethnicity
         ) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

@@

# identify number of outliers
length(
  which(
  glad.dem.raw.id.unclean$Age_raw > age_upper_limit |
    glad.dem.raw.id.unclean$Age_raw < age_lower_limit
     )
)

# remove age outliers
glad.dem.raw.id <- glad.dem.raw.id.unclean %>%
  mutate(
    Age = 
      if_else(
        Age_raw > age_upper_limit | 
        Age_raw < age_lower_limit, 
        true = NA_real_,
        false = Age_raw,
        missing = NA_real_
      )
  )

@@

# Highest education
Proceed along qualifications fields. If field is not blank, take that as highest qual. If field is blank, check next field. If all fields are blank, NA
Assumes University degree > A level > NVQ > GCSE/O Level/CSE
```{r Highest education}
#create numeric version of the highest education variable
glad.dem.raw.id <- glad.dem.raw.id %>%
  mutate(
    highest_education_numeric =
      case_when(
        dem.college_or_university_degree_numeric == "1" ~ 4,
        dem.a_levelsas_levels_or_equivalent_numeric == "1" ~ 3,
        dem.nvq_or_hnd_or_hnc_or_equivalent_numeric == "1" ~ 2,
        dem.o_levelsgcses_or_equivalent_numeric == "1" ~ 1,
        dem.cses_or_equivalent_numeric == "1" ~ 1)
    )

@@

Check highest education coding
```{r Highest education check}
glad.dem.raw.id %>%
  select(highest_education,
        dem.college_or_university_degree_numeric,
        dem.a_levelsas_levels_or_equivalent_numeric,
        dem.nvq_or_hnd_or_hnc_or_equivalent_numeric,
        dem.o_levelsgcses_or_equivalent_numeric,
        dem.cses_or_equivalent_numeric
  )
```

@@

## looking at cases over 0.5
table(dat.glad.coping$covid_odds_ratio_50) 
```

>>>>

# Select variables
```{r glad dem cleaning}
glad.dem.raw.id.unclean <- glad.dem.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
         Sample, # Sample
         ID, # ID
         startDate.glad, #Date when participant signed up to glad
         Age_uncleaned = demographics.how_old_are_you_now.txt, # Age
         Gender = demographics.which_gender_do_you_identify_with, # Gender
         Sex = demographics.select_questionnaire_items_medical, #Sex
         EduYrs = demographics.school_education_school_years.txt, # Years in education
         demographics.college_or_university_degree, # level of education - univeristy
         demographics.a_levelsas_levels_or_equivalent, # level of education - alevel
         demographics.o_levelsgcses_or_equivalent, # level of education - gcse
         demographics.cses_or_equivalent, # level of education - cse
         demographics.nvq_or_hnd_or_hnc_or_equivalent, # level of education - nvq
         Ethnicity = demographics.questions_based_ethnic_origin # Ethnicity
         ) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

Create dem numeric variables
```{r create dem numeric variables}
exclude_cols <- c("ID",
                  "Sample",
                  "startDate.glad",
                  "Age_uncleaned",
                  "EduYrs")

glad.dem.raw.id.unclean.numeric <- add_numeric(glad.dem.raw.id.unclean, exclude = exclude_cols)

# Check if factor columns have been added
colnames(glad.dem.raw.id.unclean.numeric)

# Dimensions before adding factor columns
dim(glad.dem.raw.id.unclean)
# Dimensions after adding factor columns
dim(glad.dem.raw.id.unclean.numeric)
```

@@

# identify number of outliers
length(
  which(
  glad.dem.raw.id.unclean$Age_uncleaned > age_upper_limit |
    glad.dem.raw.id.unclean$Age_uncleaned < age_lower_limit
     )
)

# remove age outliers
glad.dem.raw.id <- glad.dem.raw.id.unclean %>%
  mutate(
    Age = 
      if_else(
        Age_uncleaned > age_upper_limit | 
        Age_uncleaned < age_lower_limit, 
        true = NA_real_,
        false = Age_uncleaned,
        missing = NA_real_
      )
  )

@@

# Highest education
Proceed along qualifications fields. If field is not blank, take that as highest qual. If field is blank, check next field. If all fields are blank, NA
Assumes University degree > A level > NVQ > GCSE/O Level/CSE
```{r Highest education}
#create numeric version of the highest education variable
glad.dem.raw.id <- glad.dem.raw.id %>%
  mutate(
    highest_education_numeric =
      case_when(
        demographics.college_or_university_degree_numeric == "1" ~ 4,
        demographics.a_levelsas_levels_or_equivalent_numeric == "1" ~ 3,
        demographics.nvq_or_hnd_or_hnc_or_equivalent_numeric == "1" ~ 2,
        demographics.o_levelsgcses_or_equivalent_numeric == "1" ~ 1,
        demographics.cses_or_equivalent_numeric == "1" ~ 1)
    )

@@

Check highest education coding
```{r Highest education check}
glad.dem.raw.id %>%
  select(highest_education,
        demographics.college_or_university_degree_numeric,
        demographics.a_levelsas_levels_or_equivalent_numeric,
        demographics.nvq_or_hnd_or_hnc_or_equivalent_numeric,
        demographics.o_levelsgcses_or_equivalent_numeric,
        demographics.cses_or_equivalent_numeric
  )
```

@@

## looking at cases over 0.5
freq(dat.glad.coping$covid_odds_ratio_50, 
     cumul = FALSE) 
```
